A spectral-Galerkin turbulent channel flow solver for large-scale simulations

### 2017-08-02 paper.tex
Abstract

\begin{abstract}
A fully (pseudo-)spectral solver for direct numerical simulations of large-scale turbulent channel flows is described. The solver utilizes the Chebyshev base functions suggested by J. Shen [SIAM J. Sci. Comput., 16, 1, 1995], that lead to stable and robust numerical schemes, even at very large scale. New and fast algorithms for the direct solution of the linear systems are devised, and algorithms and matrices for all required scalar products and transforms are provided. We validate the solver for very high Reynolds numbers. Specifically, the solver is shown to reproduce the first order statistics of Hoyas and Jim\'{e}nez [Phys. Fluids, 18(1), 2006], for a channel flow at $Re_{\tau}=2000$. The solver is available through the open source project spectralDNS [https://github.com/spectralDNS].
\end{abstract}
\begin{keyword}
DNS \sep Fourier \sep Chebyshev \sep Biharmonic \sep Helmholtz \sep turbulence
\end{keyword}

@@

In this paper we will describe and assess a spectral-Galerkin channel flow solver based on Fourier and Shen's Chebyshev basis \cite{Shen95}. The solver will consist of parts that scale at worst as $N \log N$, for a 1D problem of size $N$, and as such as $N^3 \log N$ for a 3D box of size $N^3$. We will give a proper description of the theoretical basis in Sec~\ref{sec:prelim}, the discretization of Navier-Stokes equations in Sec~\ref{sec:discretizationNS}, and we will describe necessary algorithms, including a new fast direct solver for the biharmonic problem that arise, in Sec~\ref{sec:implementation}. We will finally show, in Sec~\ref{sec:verification}, that roundoff is not a major issue, and that the Shen-Fourier spectral-Galerkin method is in deed applicable to large-scale simulations. The solver has been implemented in the open source code spectralDNS \cite{spectralDNS}, where the bulk of the code is written in high-level Python \cite{python}, with critical parts migrated to Cython \cite{cython} for efficiency.

>>>>

\begin{abstract}
In this paper we describe in detail the implementation of a spectral Galerkin method intended for large scale turbulent channel flow simulations with the spectral Galerkin method. This modified Chebyshev basis was suggested some time ago by J. Shen [SIAM J. Sci. Comput., 16, 1, 1995], but has until now not been fully taken advantage of in the study of turbulence physics. What is great about this basis, that makes it ideal for the turbulent channel flow, is both its accuracy (fully spectral in all spatial directions, when combined with Fourier bases in the periodic directions) and its efficiency (all algorithms may be executed with $\mathcal{O}(N\log N)$ or better scaling). But what sets it apart from other known fully spectral channel flow solvers is the robustness and low roundoff, caused by well conditioned coefficient and inner product matrices.

For the implementation of this new solver, we devise new and fast ($\mathcal{O}(N\log N)$) algorithms for the direct solution of the linear systems, as well as algorithms and matrices for the required inner products and fast transforms. We validate the solver for very high Reynolds numbers. Specifically, the solver is shown to reproduce the first order statistics of Hoyas and Jim\'{e}nez [Phys. Fluids, 18(1), 2006], for a channel flow at $Re_{\tau}=2000$. The solver is available through the open source project spectralDNS [https://github.com/spectralDNS].

\end{abstract}
\begin{keyword}
DNS \sep Fourier \sep Chebyshev \sep Biharmonic \sep Helmholtz \sep turbulence
\end{keyword}

@@

In this paper we will describe and assess a spectral-Galerkin channel flow solver based on tensorproduct spaces of Fourier and Shen's Chebyshev basis \cite{Shen95}. The solver will consist of parts that scale at worst as $N \log N$, for a 1D problem of size $N$, and as such as $N^3 \log N$ for a 3D box of size $N^3$. We will give a proper description of the theoretical basis in Sec~\ref{sec:prelim}, the discretization of Navier-Stokes equations in Sec~\ref{sec:discretizationNS}, and we will describe necessary algorithms, including a new fast direct solver for the biharmonic problem that arise, in Sec~\ref{sec:implementation}. We will finally show, in Sec~\ref{sec:verification}, that roundoff is not a major issue, and that the Shen-Fourier spectral-Galerkin method is in deed applicable to large-scale simulations. The solver has been implemented in the open source code spectralDNS \cite{spectralDNS}, where the bulk of the code is written in high-level Python \cite{python}, with critical parts migrated to Cython \cite{cython} for efficiency.
